\mutagen: Working title

### 2021-03-30 papers/oopsla-21/main.tex
WIP

\begin{algorithm}[ b]
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{QC}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\QC{P, N, gen}}{
    i $\gets$ 0\;
    \While{i $<$ N}{
      x $\gets$ Sample(gen)\;
      \lIf{\KwNot P(x)}{\KwRet Bug(x)}
      i $\gets$ i+1 \;
    }
    \KwRet Ok\;
  }
\caption{\label{algo:quickcheck}\quickcheck Testing Loop}
\end{algorithm}

@@

To illustrate this technique, let us focus on the same motivating example used
by \citeauthor{lampropoulos2019coverage}, who propose a simple property defined over
binary trees.
%
Such data structure can be easily defined in Haskell using a custom data type
with two data constructors for leaves and branches respectively:

@@

\noindent Should we mistakenly implement \texttt{mirror}, e.g., by dropping the
right subtree altogether:

@@

@@

@@

@@

\begin{verbatim}
isValidHaskell m =
  case validate m of
    Left  validationError -> return False
    Right validModule     -> return True
\end{verbatim}

\begin{verbatim}
isValidSpec m = do
  writeFile "testcase.wasm" (dumpModule m)
  res <- shell "./wasm" ["-d", "-i", "testcase.wasm"]
  return (res == "")
\end{verbatim}

@@

\begin{verbatim}
prop_interpreter ty (args, fun) =
  discardAfter 20
  (do let m = mk_module ty "f" fun;
      resHs   <- invokeHaskell m "f" args;
      resSpec <- invokeSpec    m "f" args;
      return (equiv resHs resSpec))
\end{verbatim}

@@

\subsubsection{Real Bugs and Discrepancies found by \mutagen in \textit{haskell-wasm}}

\paragraph{Bug \#1: Invalid Memory Alignment Validation}
Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec hendrerit tempor
tellus. Donec pretium posuere tellus. Proin quam nisl, tincidunt et, mattis
eget, convallis nec, purus. Cum sociis natoque penatibus et magnis dis
parturient montes, nascetur ridiculus mus. Nulla posuere. Donec vitae dolor.
Nullam tristique diam non turpis. Cras placerat accumsan nulla. Nullam rutrum.
Nam vestibulum accumsan nisl.

\paragraph{Bug \#2: Lax invocation checks}
Aliquam erat volutpat. Nunc eleifend leo vitae magna. In id erat non orci
commodo lobortis. Proin neque massa, cursus ut, gravida ut, lobortis eget,
lacus. Sed diam. Praesent fermentum tempor tellus. Nullam tempus. Mauris ac
felis vel velit tristique imperdiet. Donec at pede. Etiam vel neque nec dui
dignissim bibendum. Vivamus id enim. Phasellus neque orci, porta a, aliquet
quis, semper a, massa. Phasellus purus. Pellentesque tristique imperdiet tortor.
Nam euismod tellus id erat.

\paragraph{Bug \#3: Allowed Out-of-bounds Memory Access}
Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

\paragraph{Discrepancy \#1: Different Reinterpretation Semantics of NaN Values}
Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec hendrerit tempor
tellus. Donec pretium posuere tellus. Proin quam nisl, tincidunt et, mattis
eget, convallis nec, purus. Cum sociis natoque penatibus et magnis dis
parturient montes, nascetur ridiculus mus. Nulla posuere. Donec vitae dolor.
Nullam tristique diam non turpis. Cras placerat accumsan nulla. Nullam rutrum.
Nam vestibulum accumsan nisl.

\paragraph{Discrepancy \#2: Allowed Blocks with Multiple Return Types}
Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec hendrerit tempor
tellus. Donec pretium posuere tellus. Proin quam nisl, tincidunt et, mattis
eget, convallis nec, purus. Cum sociis natoque penatibus et magnis dis
parturient montes, nascetur ridiculus mus. Nulla posuere. Donec vitae dolor.
Nullam tristique diam non turpis. Cras placerat accumsan nulla. Nullam rutrum.
Nam vestibulum accumsan nisl.

@@

@@

\subsection{Webassembly Engine}

@@

%% If your work has an appendix, this is the place to put it.
\appendix

\section{Detailed Empirical Results}

Fusce suscipit, wisi nec facilisis facilisis, est dui fermentum leo, quis tempor
ligula erat quis odio. Nunc porta vulputate tellus. Nunc rutrum turpis sed pede.
Sed bibendum. Aliquam posuere. Nunc aliquet, augue nec adipiscing interdum,
lacus tellus malesuada massa, quis varius mi purus non odio. Pellentesque
condimentum, magna ut suscipit hendrerit, ipsum augue ornare nulla, non luctus
diam neque sit amet urna. Curabitur vulputate vestibulum lorem. Fusce sagittis,
libero non molestie mollis, magna orci ultrices dolor, at vulputate neque nulla
lacinia eros. Sed id ligula quis est convallis tempor. Curabitur lacinia
pulvinar nibh. Nam a sapien.

>>>>

\begin{wrapfigure}{r}{0.5\textwidth}
\vspace{-10pt}
\begin{algorithm}[H]
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{QC}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\QC{P, N, gen}}{
    i $\gets$ 0\;
    \While{i $<$ N}{
      x $\gets$ Sample(gen)\;
      \lIf{\KwNot P(x)}{\KwRet Bug(x)}
      i $\gets$ i+1 \;
    }
    \KwRet Ok\;
  }
\caption{\label{algo:quickcheck}\quickcheck Testing Loop}
\end{algorithm}
\vspace{-10pt}
\end{wrapfigure}

@@

To illustrate this technique, let us focus on the same motivating example used
by \citeauthor{lampropoulos2019coverage}, who propose a simple property defined over
binary trees.
%
Such data structure can be defined in Haskell using a custom data type with two
data constructors for leaves and branches respectively:

@@

\noindent Should we mistakenly introduce a bug in \texttt{mirror}, e.g., by
dropping the right subtree altogether:

@@

We evaluated the performance of \mutagen using two main case studies.
%
In both of them, we show how the heuristics described in Section
\ref{sec:heuristics} affect the testing performance.

The first one is a simple abstract stack machine that enforces
\emph{noninterference} using runtime checks.
%
The implementation of this case study was originally proven correct by \cite{}
in Coq, and subsequently degraded by systematically introducing 20 bugs on its
enforcing mechanism.
%
\citeauthor{lampropoulos2019coverage} used this same case study to compare
\fuzzchick against random testing approaches using na\"ive and manually-written
smart random generators.
%
In this work we replicate their results and compare them against our tool.

The second case study aims to evaluate \mutagen in a more realistic scenario,
and focuses on testing \textit{haskell-wasm} an existing WebAssembly engine
written in Haskell of industrial strength.
%
We manually injected 10 bugs in the validator as well as 5 bugs in the
interpreter of this engine, and used the reference implementation to find them
using differential testing.
%
During this process, we quickly discovered 3 bugs and 2 discrepancies on this
engine with respect to the reference implementation.
%
All of them were reported and confirmed by the authors of \textit{haskell-wasm}.

@@

WebAssembly is a popular assembly-like language designed to be an open standard
for executing low-level code in web pages, although it has become increasingly
popular in standalone, non-web contexts as well.
%
WebAssembly programs are first validated and later executed in a sandboxed
environment (a virtual machine), making this language an attractive target for
virtualization in \emph{functions-as-a-service} platforms.
%
The language is relatively simple, in esence
%
\begin{inparaenum}
  \item it contains only four numerical types, representing both integers and
    IEEE754 floating point numbers of either 32 or 64 bits;
  \item values of these types are manipulated by writing functions using
    sequences of stack instructions;
  \item functions are organized in modules and must be explicitly imported and
    exported;
  \item memory blocks can be imported, exported and grown dinamically;
\end{inparaenum}
%
among others.
%
To give the reader a taste, Fig. \ref{fig:wasm:factorial} defines a WebAssembly
module that exports a recursive implementation of the factorial function.
%
This language is a perfect match for our tool, and the space of WebAssembly
programs contains mostly invalid ones, and automatically derived random
generators cannot satisfy all the invariants required to produce interesting
ones.

\begin{wrapfigure}{r}{0.35\textwidth}
\vspace{-10pt}
\begin{verbatim}
(module
(func $f (param f64) (result f64)
  get_local 0
  f64.const 1
  f64.lt
  if (result f64)
    f64.const 1
  else
    get_local 0
    get_local 0
    f64.const 1
    f64.sub
    call $fact
    f64.mul
  end)
(export "fact" (func $f)))
\end{verbatim}
\caption{\label{fig:wasm:factorial}Factorial in WebAssembly}
\vspace{-10pt}
\end{wrapfigure}

Unlike most other programming languages, its behavior is fully specified, and
WebAssembly programs are expected to be consistently interpreted across engines
--- despite some subtle details that we will address soon.
%
For this purpose, the WebAssembly standard provides a reference engine
implementation with all the basic functionality expected from a compliant
WebAssembly engine.

In this work, we are interested on using our tool to test the two most complex
subsystems of \textit{haskell-wasm}: the \emph{validator} and the
\emph{interpreter} --- both being previously tested using a unit test suite.
%
For this purpose, we avoided spending countless hours writing an extensive
property-based specification to mimic the reference WebAssembly specification.
%
Instead, we take advantage of the readily available reference implementation via
differential testing.
%
In this light, our testing properties assert that any result produced by
\textit{haskell-wasm} matches that of the reference implementation.

Unsurprisingly, this engine had several subtle latent bugs that were not caught
by the existing unit tests and that we discovered using \mutagen while
developing the test suite used in this work.
%
Moreover, \mutagen exposed two discrepancies between \textit{haskell-wasm} and
the reference implementation.
%
Not severe enough to be classified as bugs, these discrepancies trigger parts of
the WebAssembly specification that are either not yet suported by the reference
implementation (multi-value blocks), or that produce a well-known
non-deterministic undefined behavior allowed by the specification (NaN
reinterpretation).
%
These findings are briefly outlined in Table \ref{table:wasm:bugs}.

The rest of this section introduces the test suite we used to test the different
parts of \textit{haskell-wasm}.

\begin{table}[b]
\footnotesize
\begin{tabular}{|c|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Category}
& \textbf{Description} \\
\hline
1
& Validator
& Bug
& Invalid memory alignment validation \\
\hline
2
& Validator
& Discrepancy
& Validator accepts blocks returning multiple values  \\
\hline
3
& Interpreter
& Bug
& Instance function invoker silently proceeds after arity mismatch \\
\hline
4
& Interpreter
& Bug
& Allowed out-of-bounds memory access \\
\hline
5
& Interpreter
& Discrepancy
& NaN reinterpretation does not follow reference implementation\\
\hline
\end{tabular}
\caption{\label{table:wasm:bugs}Bugs and discrepancies found by \mutagen in
  \textit{haskell-wasm}.}
\end{table}

@@

We begin by designing a property to test the WebAssembly validator implemented
in \textit{haskell-wasm}.
%
To keep things simple, we can simply assert that, whenever a randomly generated
(or mutated for that matter) WebAssembly module is valid according to
\textit{haskell-wasm}, then the reference implementation agrees upon it.
%
In other words, we are testing for false positives.
%
In Haskell, we write the following predicate:

@@

Where the precondition (\texttt{isValidHaskell m}) runs the input WebAssembly
module \texttt{m} against the \textit{haskell-wasm} validator, whereas the
postcondition (\texttt{isValidSpec m}) serializes \texttt{m} to a file, runs it
against the reference implementation validator and checks that no errors are
produced.

We want to remark that, although here for simplicity we only focus on finding
false positives, in a realistic test suite, one would also want to test for
false negatives, i.e., when a module is valid and \textit{haskell-wasm} rejects
it.
%
This can be easily done by inverting the direction of the implication
(\texttt{<==}) in the property above.
%
However, the resulting property is much slower, as every tested module will
first be serialized and run against the reference implementation.

%% \begin{verbatim}
%% isValidHaskell m =
%%   case validate m of
%%     Left  validationError -> return False
%%     Right validModule     -> return True
%% \end{verbatim}

%% \begin{verbatim}
%% isValidSpec m = do
%%   writeFile "testcase.wasm" (dumpModule m)
%%   res <- shell "./wasm" ["-d", "-i", "testcase.wasm"]
%%   return (res == "")
%% \end{verbatim}

@@

\begin{verbatim}
prop_interpreter ty (args, fun) =
  discardAfter 20
  (do let m = mk_module ty "f" fun;
      resHs   <- invokeHaskell m "f" args
      resSpec <- invokeSpec    m "f" args
      return (equivalent resHs resSpec))
\end{verbatim}

@@

%% \subsubsection{Real Bugs and Discrepancies found by \mutagen in \textit{haskell-wasm}}

%% \paragraph{Bug \#1: Invalid Memory Alignment Validation}
%% \paragraph{Bug \#2: Lax invocation checks}
%% \paragraph{Bug \#3: Allowed Out-of-bounds Memory Access}
%% \paragraph{Discrepancy \#1: Different Reinterpretation Semantics of NaN Values}
%% \paragraph{Discrepancy \#2: Allowed Blocks with Multiple Return Types}

@@

All the experiments were done in a dedicated workstation with an Intel Core
i7-8700 CPU running at 3.20GHz, and equipped with 32Gb of RAM.
%
We ran each experiment 30 times except for the ones involving the bugs on the
WebAssembly interpreter, which were run 10 times.
%
From there, we followed the same approach taken by
\citeauthor{lampropoulos2019coverage} and collected the Mean-Time-To-Failure
(MTTF) of each bug, i.e., how quickly a bug can be found in wall time.
%
In all cases, we used a one-hour timeout to stop the execution of both tools if
they have not yet found a counterexample.

Additionally, we collected the Failure Rate (FR) observed for each bug, i.e. the
proportion of times each tool finds each bug within the one-hour testing bugdet.
%
We found this metric crucial to be analyzed when replicating \fuzzchick's
results, as opposed to just paying attention at the MTTF.

@@

@@

%% If your work has an appendix, this is the place to put it.
%% \appendix
